---
title: "covid19_spread"
output: html_document
date: "2025-11-15"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
library(dplyr)
library(knitr)
```

### Team members

-   **Bohdan Zasymovych**

-   **Solomiia Gadiichuk**

-   **Iryna Svyrydenko**

## Statistical Analysis of COVID-19 Spread

### The aim of this project is to analyze open data about Covid-19.

For this we are using the dataset from the collection of the COVID-19 data maintained by Our World in Data (<https://github.com/owid/covid-19-data/tree/master>). The data is gathered from official resources (national governments and ministries of Health, worldwide organizations (UN, WHO)).

We decided to investigate the connection between vaccination and mortality.

For this we are using such random variables:

-   `new_cases_smoothed_per_million` - new confirmed cases of COVID-19 (7-day smoothed) per 1,000,000 people;
-   `new_deaths_smoothed_per_million` - new deaths attributed to COVID-19 (7-day smoothed) per 1,000,000 people;
-   `people_fully_vaccinated_per_hundred` - total number of people who received all doses prescribed by the initial vaccination protocol per 100 people in the total population;
-   `stringency_index` - Government Response Stringency Index: composite measure based on 9 response indicators including school closures, workplace closures, and travel bans, rescaled to a value from 0 to 100 (100 = strictest response).

```{r}
{df <- read.csv("owid-covid-data.csv")}

```

```{r}
selected_countries <- c("World", "Ukraine", "United States",
                        "Brazil", "Australia", "France")

df_filtered <- df[df$location %in% selected_countries, ]

df_small <- df_filtered[, c(
  "date", 
  "location",
  "new_cases_smoothed_per_million",
  "new_deaths_smoothed_per_million",
  "people_fully_vaccinated_per_hundred",
  "stringency_index"
)]
df_small$date <- as.Date(df_small$date)
kable(head(df_small, 10), caption = "Preview of Filtered COVID-19 Data")

```

```{r, fig.width=14, fig.height=10}
ggplot(df_small, aes(x = date,
                        y = new_cases_smoothed_per_million,
                        color = location)) +
  geom_line(linewidth = 1, na.rm = TRUE) +
    scale_color_manual(values = c(
    "Australia" = "#E41A1C",      
    "Brazil" = "#377EB8",         
    "France" = "#4DAF4A",         
    "Ukraine" = "#984EA3",        
    "United States" = "#FF7F00",  
    "World" = "black"             
  )) +
  
  facet_wrap(~location, scales = "free_y", ncol = 2) +
  theme_minimal(base_size = 16) +
  labs(title = "COVID-19 Case Trends by Country",
       x = "Date",
       y = "New Cases per Million",
       color = "Country") +
  theme(legend.position = "none")
```

This graph shows change of the `new_cases_smoothed_per_milion` random variable with time for 5 chosen country and a world.

```{r, fig.width=14, fig.height=8}
ggplot(df_small, aes(x = date,
                     y = people_fully_vaccinated_per_hundred,
                     color = location)) +
  geom_line(size = 1.2, na.rm = TRUE) +
  scale_color_brewer(palette = "Set1") +
  theme_minimal(base_size = 16) +
  labs(title = "Vaccination Progress in Selected Countries",
       x = "Date",
       y = "Fully Vaccinated per Hundred",
       color = "Country") +
  theme(
    plot.title = element_text(face = "bold", size = 20),
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
```

This graph shows change of the `people_fully_vaccinated_per_hundred` random variable with time for 5 chosen country and a world.

```{r, fig.width=16, fig.height=12}
ggplot(df_small, aes(x = date,
                     y = new_deaths_smoothed_per_million,
                     color = location)) +
  geom_line(size = 1, na.rm = TRUE) +
  facet_wrap(~ location, scales = "free_y", ncol = 2) +
  theme_minimal(base_size = 16) +
    scale_color_manual(values = c(
    "Australia"      = "#E41A1C",   
    "Brazil"         = "#377EB8",  
    "France"         = "#4DAF4A",   
    "Ukraine"        = "#984EA3",   
    "United States"  = "#FF7F00",   
    "World"          = "black"     
  )) +
  labs(
    title = "COVID-19 Death Trends by Country",
    x = "Date",
    y = "New Deaths per Million"
  ) +
  theme(
    legend.position = "none",
    plot.title = element_text(face = "bold", size = 20),
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
```

This graph shows change of the `new_deaths_smoothed_per_million` random variable with time for 5 chosen country and a world.

```{r}
date <- "2021-08-01"
target_date <- as.Date(date)

snapshot <- df %>%
  filter(date == target_date) %>%
  filter(!is.na(people_fully_vaccinated_per_hundred),
         !is.na(new_deaths_smoothed_per_million),
         !location %in% c("World")
  ) %>%
  select(location,
         people_fully_vaccinated_per_hundred,
         new_deaths_smoothed_per_million)

cat("Number of countries with available data for this date: ", length(df))

ggplot(snapshot,
       aes(x = people_fully_vaccinated_per_hundred,
           y = new_deaths_smoothed_per_million)) +

  geom_point(alpha = 0.6, size = 2) +
  labs(
    title = paste("Vaccination vs COVID-19 Death Rate on", target_date),
    x = "Fully Vaccinated per Hundred",
    y = "New Deaths per Million"
  ) +
  theme_minimal()
```

On this graph we tried to get initial insight about impact of vaccines on a death rate. Each point represents a country on a specific date (August 1, 2021). The X-axis shows the percentage of the population fully vaccinated `people_fully_vaccinated_per_hundred`, while the Y-axis shows the smoothed number of new deaths per million people `new_deaths_smoothed_per_million`.

```{r}
df_hypothesis2 <- df %>%
  filter(!is.na(aged_65_older), 
         !is.na(new_deaths_smoothed_per_million),
         !is.na(continent), continent != "") %>%
  group_by(location) %>%
  summarise(
    aged_65_older = max(aged_65_older, na.rm = TRUE),
    total_deaths_per_million = max(total_deaths_per_million, na.rm = TRUE)
  )

ggplot(df_hypothesis2, aes(x = aged_65_older, y = total_deaths_per_million)) +
  geom_point(alpha = 0.6, color = "darkgreen") +
  labs(
    title = "Relationships between population age and Covid-19 mortality",
    x = "Percentage of people over the age of 65",
    y = "New deaths smoothed per million"
  ) +
  theme_minimal()
```

In this graph the X-axis shows the percentage of the people over the age of 65 `aged_65_older`, and Y-axis shows the smoothed number of new deaths per million people `new_deaths_smoothed_per_million`.

# Hypotheses

### Our first hypothesis

From the graph "Vaccination vs COVID-19 Death Rate on 2021-08-01", we observed that countries with a higher percentage of fully vaccinated people tend to have points closer to the X-axis. Based on this observation, we formulated the following hypothesis to test whether this pattern occurred by chance or there is general trend.

$H_0$: Vaccination rate has no effect on death rate.

$H_1$: Increasing vaccination rate reduces the death rate.

### Our second hypothesis

In the graph "Relationships between population age and Covid-19 mortality", we noticed that countries with a higher percentage of people over 65 tend to have points further from the X-axis. Based on this observation, we formulated the following hypothesis to test whether this pattern occurred by chance or there is general trend.

$H_0$: Countries with a higher proportion of people over 65 have higher COVID-19 death rates..

$H_1$: Countries with a higher proportion of people over 65 have lower COVID-19 death rates..
