---
title: "covid19_spread"
output: html_document
date: "2025-11-15"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
```

## Statistical Analysis of COVID-19 Spread

### The aim of this project is to analyze open data about Covid-19.

For this we are using the dataset from the collection of the COVID-19 data maintained by Our World in Data (<https://github.com/owid/covid-19-data/tree/master>). The data is gathered from official resources (national governments and ministries of Health, worldwide organizations (UN, WHO)).

We decided to investigate the connection between vaccination and mortality.

For this we are using such random variables:

-   **new_cases_smoothed_per_million** - new confirmed cases of COVID-19 (7-day smoothed) per 1,000,000 people;
-   **new_deaths_smoothed_per_million** - new deaths attributed to COVID-19 (7-day smoothed) per 1,000,000 people;
-   **people_fully_vaccinated_per_hundred** - total number of people who received all doses prescribed by the initial vaccination protocol per 100 people in the total population;
-   **stringency_index** - Government Response Stringency Index: composite measure based on 9 response indicators including school closures, workplace closures, and travel bans, rescaled to a value from 0 to 100 (100 = strictest response).

```{r}
{df <- read.csv("owid-covid-data.csv")}
df
```

```{r}
selected_countries <- c("World", "Ukraine", "United States",
                        "Brazil", "Australia", "France")

df_filtered <- df[df$location %in% selected_countries, ]

df_small <- df_filtered[, c(
  "date", 
  "location",
  "new_cases_smoothed_per_million",
  "new_deaths_smoothed_per_million",
  "people_fully_vaccinated_per_hundred",
  "stringency_index"
)]
df_small$date <- as.Date(df_small$date)
head(df_small)

```

```{r}
colSums(is.na(df_small))
```

```{r, fig.width=14, fig.height=10}
ggplot(df_small, aes(x = date,
                        y = new_cases_smoothed_per_million,
                        color = location)) +
  geom_line(size = 1, na.rm = TRUE) +
  scale_color_brewer(palette = "Set1") +
  facet_wrap(~location, scales = "free_y", ncol = 2) +
  theme_minimal(base_size = 16) +
  labs(title = "COVID-19 Case Trends by Country",
       x = "Date",
       y = "New Cases per Million",
       color = "Country") +
  theme(legend.position = "none")
```

```{r, fig.width=14, fig.height=8}
ggplot(df_small, aes(x = date,
                     y = people_fully_vaccinated_per_hundred,
                     color = location)) +
  geom_line(size = 1.2, na.rm = TRUE) +
  scale_color_brewer(palette = "Set1") +
  theme_minimal(base_size = 16) +
  labs(title = "Vaccination Progress in Selected Countries",
       x = "Date",
       y = "Fully Vaccinated per Hundred",
       color = "Country") +
  theme(
    plot.title = element_text(face = "bold", size = 20),
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
```

```{r, fig.width=16, fig.height=12}
ggplot(df_small, aes(x = date,
                     y = new_deaths_smoothed_per_million,
                     color = location)) +
  geom_line(size = 1, na.rm = TRUE) +
  facet_wrap(~ location, scales = "free_y", ncol = 2) +
  theme_minimal(base_size = 16) +
  scale_color_brewer(palette = "Set1") +
  labs(
    title = "COVID-19 Death Trends by Country",
    x = "Date",
    y = "New Deaths per Million"
  ) +
  theme(
    legend.position = "none",
    plot.title = element_text(face = "bold", size = 20),
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
```
